Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-15 main.tex
Cut out algorithm 1 and change the explanation style a bit (close #11)

\begin{wrapfigure}{r}{0.5\textwidth}
\vspace{-20pt}
\begin{minipage}{0.5\textwidth}
\begin{algorithm}[H]
\caption{Simple translation algorithm for the POM-Parse Tree}\label{alg:simple-translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of the parse tree $T$
	\Procedure{translate}{$r$}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$);
	\Else
		\ForAll{children $v_n$ of $r$}
			\State {\scriptsize TRANSLATE}($v_n$);
		\EndFor
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\end{minipage}
\vspace{-14pt}
\end{wrapfigure}

@@

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. The Algorithm~\ref{alg:simple-translation} explains the translation process. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively.

\begin{algorithm}[t]
\caption{Abstract translation algorithm to translate MLP-Parse trees.}\label{alg:translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of a POM-Parse tree $T$. List \textit{following\_siblings} with the following siblings of $r$. The list can be empty.
	\Procedure{abstract\_translator}{$r$, \textit{following\_siblings}}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$, \textit{following\_siblings});\label{line:transLeaf}
	\Else
		\State \textit{siblings} $ = r$.getChildren(); \Comment{\textit{siblings} is a list of children}
		\State {\scriptsize ABSTRACT\_TRANSLATOR}(\textit{siblings}.removeFirst(), \textit{siblings});\label{line:transSeq}
	\EndIf
	\If{\textit{following\_siblings} is not empty}\label{line:ifNotEmpty}
		\State $r =$ \textit{following\_siblings}.removeFirst();
		\State {\scriptsize ABSTRACT\_TRANSLATOR}($r$, \textit{following\_siblings});
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

This approach is not completely feasible since the algorithm needs to look ahead and check the following siblings in some cases, e.g., in the case of a semantic macro with arguments (see~\Cref{fig:syntax-tree-usecase}). The {\footnotesize \verb|TRANSLATE_LEAF|} function needs to know the following siblings of the current node. Algorithm~\ref{alg:translation} explains a more detailed but abstract version of the final algorithm of the presented translation tool.

@@

\begin{table}[t!]
\centering
\sloppy
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{Y}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}
\begin{tabularx}{\textwidth}{Y{1.55cm} Y{2.5cm} L L}
	\hline
	& \textbf{Node type} & \textbf{Explanation} & \textbf{Example} \\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ has children}} & Sequence & Contains a list of expressions. & $a+b$ is a sequence with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Balanced Expression & Similar to a sequence. But in this case the sequence is wrapped by \texttt{\tbs left} and \texttt{\tbs right} delimiters. & \texttt{\tbs left(}$a+b$\texttt{\tbs right)} is a balanced expression with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Fraction & All kinds of fractions, such as \texttt{\tbs frac}, \texttt{\tbs ifrac}, etc. & \texttt{\tbs ifrac\{a\}\{b\}} is a fraction with two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Binomial & Binomials & \texttt{\tbs binom\{a\}\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Square Root & The square root with one child. & \texttt{\tbs sqrt\{a\}} has one child ($a$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Radical with a specified index & $n$-th root with two children. & \texttt{\tbs sqrt[a]\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Underscore & The underscore `\_' for subscripts. & The sequence $a$\_$b$ has two children ($a$ and `\_'). The underscore itself `\_' has one child ($b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Caret & The caret `\^{}' for superscripts or exponents. Similar to the underscore. & The sequence $a$\^{}$b$ has two children ($a$ and `\^{}'). The caret itself `\^{}' has one child ($b$).\\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ is a leaf}} & \Macro{} & A semantic \LaTeX{} macro & \texttt{\tbs JacobiP}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Generic \LaTeX{} macro & All kinds of \LaTeX{} macros & \texttt{\tbs Rightarrow}, \texttt{\tbs alpha}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Alphanumerical Expressions & Letters, numbers and general strings. & Depends on the order of symbols. $ab3$ is alphanumerical, while $4b$ are two nodes ($4$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Symbols & All kind of symbols & `$@$', `$*$', `$+$', `$!$', etc.\\
	\hline
\end{tabularx}
\caption{A table of all kinds of nodes in a PoM syntax tree. Note that this table groups some types together for a better overview. For a complete list and a more detailed version see~\parencite{POM-Tagger}.}
\label{tab:allTypesTable}
\end{table} 
This approach seems to work for most semantic \LaTeX{} expressions. However, there is one problem with this straightforward approach which was mentioned in~\Cref{sec:problems}. There are many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the simple expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense as long as the operator is a binary operator.

@@

Most programming languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, since it is more convenient to parse expressions in uniform notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules yet, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with most common notations. Therefore, the translator does not need to know that `$a$' and `$b$' are the operands of the binary operator `$+$' in `$a+b$.' The translator could simply translate the symbols in `$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, this simple approach generates two problems.
\begin{enumerate}\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to \gls*{cas} which use non-standard notations.
\end{enumerate}

Problem~\ref{prob:1} should be obvious. Since we want to develop a translation tool and not a verification tool for mathematical \LaTeX{} expressions, we can assume syntactically correct input expressions and produce errors otherwise. Problem~\ref{prob:2} is more difficult to solve. If a user wants to support a \gls*{cas} that uses prefix or postfix notation by default, the translator would fail in its current state. Supporting \gls*{cas} with another notation would be a part of future work.

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguates the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example is already interpreted as the double factorial function of $n$, since this notation is the standard interpretation in science. We wrote the second interpretation as the standard way in science to make it even more obvious. However, surprisingly, \Maple{} computes the first interpretation again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}

>>>>

@@

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively. This idea seems to be a practical and elegant solution but it has a significant drawback, it cannot be used to translate functions. Since the arguments of functions are following siblings in the \gls*{pom-pt}, the algorithm needs to look ahead when a leaf is a known function, e.g., in the case of a semantic macro with arguments (see~\Cref{fig:syntax-tree-usecase}). Algorithm~\ref{alg:translation} is an improved version with look ahead functionality.

\begin{algorithm}[t]
\caption{Abstract translation algorithm to translate PPT.}\label{alg:translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of a PoM-Parse tree $T$. List \textit{following\_siblings} with the following siblings of $r$. The list can be empty.
	\Procedure{abstract\_translator}{$r$, \textit{following\_siblings}}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$, \textit{following\_siblings});\label{line:transLeaf}
	\Else
		\State \textit{children} $ = r$.getChildren(); \Comment{\textit{children} is a list of nodes}
		\State {\scriptsize ABSTRACT\_TRANSLATOR}(\textit{children}.removeFirst(), \textit{children});\label{line:transSeq}
	\EndIf
	\If{\textit{following\_siblings} is not empty}\label{line:ifNotEmpty}
		\State $r =$ \textit{following\_siblings}.removeFirst();
		\State {\scriptsize ABSTRACT\_TRANSLATOR}($r$, \textit{following\_siblings});
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

@@

\begin{table}[t!]
\centering
\sloppy
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{Y}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}
\begin{tabularx}{\textwidth}{Y{1.55cm} Y{2.5cm} L L}
	\hline
	& \textbf{Node type} & \textbf{Explanation} & \textbf{Example} \\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ has children}} & Sequence & Contains a list of expressions. & $a+b$ is a sequence with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Balanced Expression & Similar to a sequence. But in this case the sequence is wrapped by \texttt{\tbs left} and \texttt{\tbs right} delimiters. & \texttt{\tbs left(}$a+b$\texttt{\tbs right)} is a balanced expression with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Fraction & All kinds of fractions, such as \texttt{\tbs frac}, \texttt{\tbs ifrac}, etc. & \texttt{\tbs ifrac\{a\}\{b\}} is a fraction with two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Binomial & Binomials & \texttt{\tbs binom\{a\}\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Square Root & The square root with one child. & \texttt{\tbs sqrt\{a\}} has one child ($a$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Radical with a specified index & $n$-th root with two children. & \texttt{\tbs sqrt[a]\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Underscore & The underscore `\_' for subscripts. & The sequence $a$\_$b$ has two children ($a$ and `\_'). The underscore itself `\_' has one child ($b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Caret & The caret `\^{}' for superscripts or exponents. Similar to the underscore. & The sequence $a$\^{}$b$ has two children ($a$ and `\^{}'). The caret itself `\^{}' has one child ($b$).\\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ is a leaf}} & \Macro{} & A semantic \LaTeX{} macro & \texttt{\tbs JacobiP}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Generic \LaTeX{} macro & All kinds of \LaTeX{} macros & \texttt{\tbs Rightarrow}, \texttt{\tbs alpha}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Alphanumerical Expressions & Letters, numbers and general strings. & Depends on the order of symbols. $ab3$ is alphanumerical, while $4b$ are two nodes ($4$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Symbols & All kind of symbols & `$@$', `$*$', `$+$', `$!$', etc.\\
	\hline
\end{tabularx}
\caption{A table of all kinds of nodes in a PoM syntax tree. Note that this table groups some types together for a better overview. For a complete list and a more detailed version see~\parencite{POM-Tagger}.}
\label{tab:allTypesTable}
\end{table}

The look ahead functionality seems to solve the problems for functions. But there is another problem that was not mentioned in \Cref{sec:problems}, because it seems to be not a problem previously. In some cases the arguments of a function do not follow but precede the function node.

If we take a closer look to mathematical notations we discover many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense as long as the operator is a binary operator.

@@

Most programming languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, since it is more convenient to parse expressions in uniform notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules yet, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from~\Cref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with most common notations. Therefore, the translator does not need to know that `$a$' and `$b$' are the operands of the binary operator `$+$' in `$a+b$.' The translator could simply translate the symbols in `$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, there are two new problems with this approach.
\begin{enumerate}\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to \gls*{cas} which use non-standard notations.
\end{enumerate}

Problem~\ref{prob:1} should be obvious. Since we want to develop a translation tool and not a verification tool for mathematical \LaTeX{} expressions, we can assume syntactically correct input expressions and produce errors otherwise. Problem~\ref{prob:2} is more complex. If a user wants to support a \gls*{cas} that uses prefix or postfix notation by default, the translator would fail in its current state. Supporting \gls*{cas} with another notation would be a part of future work.

Nonetheless, adopting different notations, in some situations, could also solve potential ambiguities. Consider the two potentially ambiguous examples in~\Cref{tab:amb_ex}. While a scientist would probably just ask for the right interpretation of the first example, \Maple{} automatically computes the first interpretation. On the other hand, \LaTeX{} automatically disambiguates the first example by only recognizing the very next element (single symbols or sequences in curly brackets) for the superscript and therefore displays the second interpretation. The second example should not be misinterpreted, since this notation is the standard interpretation in science for the double factorial. We wrote the second interpretation with parentheses point it out more precisely. However, surprisingly, \Maple{} computes the first interpretation (the factorial of the factorial of $n$) again rather than the common standard interpretation.
\begin{table}[ht]
\centering
\begin{tabular}{lccc}
	\hline
	& Text Format Expression & First Interpretation & Second Interpretation\\
	\hline
	1: & \rule{0pt}{0.9\normalbaselineskip} $4\ \hat{\ }\ 2!$ & $4^{2!}$ & $4^2!$ \\
	2: & $n!!$ & $(n!)!$ & $(n)!!$\\
	\hline
\end{tabular}
\caption{Potentially ambiguous examples of the factorial and double factorial function. One expression in a text format can be interpreted in different ways.}
\label{tab:amb_ex}
\end{table}
